---
title: "_Palabras que se las lleva el viento_"
subtitle: "Exploring presidential inauguration speeches of the last decade"
output: html_document
---

## Introduction

This is a project to explore the priorities of Peruvian governments in the last decade. The data used is the presidential inauguration speeches of the last decade. The speeches are available in the website of the Peruvian Congress. The speeches are in Spanish, so I used Google Translate to translate them to English. The speeches are available in the folder "data".

## Data

```{r libraries}
library(tidyverse)

```

```{r download speeches}
presidents <-
c("Dina Boluarte", 
  "Pedro Castillo", 
  "Francisco Sagasti", 
  "Martin Vizcarra", 
  "Pedro Pablo Kuczynski", 
  "Ollanta Humala", 
  "Alejandro Toledo", 
  "Valentín Paniagua", 
  "Alberto Fujimori second term", 
  "Alberto Fujimori first term")

presidents <- tolower(gsub(" ", "_", presidents))

links_github <- paste0(
  "https://raw.githubusercontent.com/carolinacornejocastellano/presidential_text_mining/main/",
  presidents, ".md")
```

```{r}
# Read speeches
speeches$speech <- speeches$url %>% 
    map_chr(~ read_lines(.x, locale = locale(encoding = "UTF-8")) %>% 
    paste(collapse = "\n"))
```



```{r}
# Load speeches from GitHub
speeches <- tibble(
  president = c("Dina Boluarte", "Pedro Castillo", "Francisco Sagasti", "Martin Vizcarra", "Pedro Pablo Kuckzynski", "Ollanta Humala", "Alejandro Toledo", "Valentín Paniagua", "Alberto Fujimori (second term)", "Alberto Fujimori (first term)"),
  url = c("https://raw.githubusercontent.com/carolinacornejocastellano/presidential_text_mining/main/dina_boluarte.md",
          "https://raw.githubusercontent.com/carolinacornejocastellano/presidential_text_mining/main/pedro_castillo.md",
          "https://raw.githubusercontent.com/carolinacornejocastellano/presidential_text_mining/main/francisco_sagasti.md",
          "https://raw.githubusercontent.com/carolinacornejocastellano/presidential_text_mining/main/martin_vizcarra.md",
          "https://raw.githubusercontent.com/carolinacornejocastellano/presidential_text_mining/main/pedro_pablo_kuczynski.md",
          "https://raw.githubusercontent.com/carolinacornejocastellano/presidential_text_mining/main/ollanta_humala.md",
          "https://raw.githubusercontent.com/carolinacornejocastellano/presidential_text_mining/main/alejandro_toledo.md",
          "https://raw.githubusercontent.com/carolinacornejocastellano/presidential_text_mining/main/valentin_paniagua.md",
          "https://raw.githubusercontent.com/carolinacornejocastellano/presidential_text_mining/main/alberto_fujimori_second_term.md",
          "https://raw.githubusercontent.com/carolinacornejocastellano/presidential_text_mining/main/alberto_fujimori_first_term.md")
)

# Read speeches
speeches$speech <- speeches$url %>% map_chr(~ read_lines(.x, locale = locale(encoding = "UTF-8")) %>% paste(collapse = "\n"))
```



## N-grams 

## Sentiment analysis

## Latent Dirichlet Allocation
augment() function from the tidytext package to add a column with the number of words in each speech.

```{r}